\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-13 paper.tex
Minor updates

% FIXME abstract needs some rewriting, we're placing too much emphasis
% on the coalescent there.
\begin{abstract}
New developments have made it possible to infer genetic genealogies in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
The structure representing such recombinant genetic ancestry
is usually referred to as an ancestral recombination graph (ARG),
although there is some confusion about the interpretation and little
agreement on specific details.
We propose a concrete definition of ARGs
in terms of genomes and their intervals of genetic inheritance (gARG)
and contrast this with the classical event-based definition (eARG).
We show that eARGs are limited in the patterns
of genetic inheritance that can be represented,
and require complete precision about all recombination events.
We show, in contrast, that the gARG definition
can fully capture the richness of modern large-scale datasets,
enables fine-grained
levels of precision about recombination to be represented,
and forms the basis of an efficient computational framework.
\end{abstract}

@@

The word ``ARG'' is now widely used to refer to genetic genealogies in
general~\citep[][AND OTHERS]{mathieson2020ancestry}, including
varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
This broad usage means there is a lack of consensus on the
exact object being defined. This causes problems, for example,
when evaluating inferred ARGs, and even when defining
what the criteria for success should be.
It also means there is no shared format for interchanging ARG data between
programs, resulting in the assorted negative consequences with which
geneticists are only too familiar~\citep{excoffier2006computer,gower2022demes}.
Perhaps most importantly, the direct linkage to
the original stochastic process definition leads to an
overly narrow interpretation of ARGs, which is significantly
out of step with modern methods and large-scale datasets, and
can lead to missed opportunities in their application and usage.
Such problems go much deeper than pedantic discussion over terminology;
they directly inform what we should be attempting to estimate, how we
assess success, and the expected limitations on the precision
and accuracy of our inferences.
% Thanks Gideon and Alex!
If not addressed, these problems will
impede progress in applying genetic genealogies to understand the biological world \citep{lewanski2023era}.

@@

Fig.~\ref{fig-ancestry-resolution}C shows the graph that we obtain
by resolving the ancestral material transmitted along each edge
using the ``simplify'' algorithm~\citep{kelleher2018efficient},
and highlights some important
differences between the eARG and gARG encodings.
Firstly, we can see that some nodes and edges are removed entirely
from the graph by this conversion process.
The ``grand MRCA'' \noderef{q} is omitted from the
sample-resolved gARG because all segments of the genome have
fully coalesced before it is reached. Likewise, the edge
between \noderef{g} and \noderef{j} is omitted because the recombination
event at position $5$ (represented by node \noderef{g})
fell in non-ancestral material.
More generally, we can see that the sample resolved
retrospective gARG of Fig.~\ref{fig-ancestry-resolution}C
allows for ``local'' inspection
of an ARG in a way that is not possible by storing
common ancestor and recombination events in an eARG. Because
the ancestral material is stored with each edge, the
cumulative effects of events over time can be reasoned
about, without first ``replaying'' those events. Many computations
that we wish to perform on an ARG will require resolving
the ancestral material with respect to a sample.
% this is weak, but would be good to emphasise this point.
% you just end up running simplify lots of times, in your
% downstream algs
The gARG encoding
allows us to do this X once and to store the result, whereas
the eARG encoding requires us to repeat the process of X
each time.

@@

The gARG encoding discussed and concretely defined in this manuscript
leads to highly efficient storage and processing of ARG data,
and has already been in use for several years.
The succinct tree sequence data structure (usually known as a ``tree sequence''
for brevity)
is a concrete encoding of a gARG, as discussed in this article.
It was originally developed as part of the \texttt{msprime}
simulator~\citep{kelleher2016efficient} and has subsequently been
extended and applied to forward-time
simulations~\citep{kelleher2018efficient,haller2018tree},
inference from data~\citep{kelleher2019inferring,wohns2022unified},
and calculation of population genetics statistics~\citep{ralph2020efficiently}.
The succinct tree sequence encoding extends the basic definition
of a gARG provided here by stipulating a
simple tabular representation and nodes and edges,
and also defining a concise and lossless representation of
sequence variation using the ``site'' and  ``mutation'' tables.
(Technically, edges in tree sequence terminology would be better
described as ``edge-intervals'', as each describes a single contiguous
interval of genome inheritance between a pair of nodes. This
denormalisation of the gARG data model is for efficiency purposes.)
The \texttt{tskit} library is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with gARGs (encoded as a succinct tree sequence).
Based on core functionality written
in C, it provides interfaces in C, Python and Rust.
Tskit is mature software, widely used in population genetics, and
has been incorporated into several downstream
applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
adrion2020community,
terasaki2021geonomics,
baumdicker2021efficient,
fan2022genealogical,korfmann2022weak,
mahmoudi2022bayesian,petr2022slendr,rasmussen2022espalier,
zhang2023biobank,nowbandegani2023extremely,ignatieva2023distribution}.

>>>>

% FIXME abstract needs some rewriting, we're placing too much emphasis
% on the coalescent there.
\begin{abstract}
New developments have made it possible to infer genetic genealogies in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
The structure representing such recombinant genetic ancestry
is usually referred to as an ancestral recombination graph (ARG),
although there is some confusion about the interpretation and little
agreement on specific details.
We propose a concrete definition of ARGs
in terms of genomes and their intervals of genetic inheritance (gARG)
and contrast this with the classical event-based definition (eARG).
We show that eARGs are limited in the patterns
of genetic inheritance that can be represented,
and require a precise description of the details of all recombination events.
We show, in contrast, that the gARG definition
can fully capture the richness of modern large-scale datasets,
enables fine-grained
levels of precision about recombination to be represented,
and forms the basis of an efficient computational framework.
\end{abstract}

@@

The word ``ARG'' is now widely used to refer to genetic genealogies in
general~\citep[][AND OTHERS]{mathieson2020ancestry}, including
varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
This broad usage means there is a lack of consensus on the
exact object being defined. This causes problems, for example,
when evaluating inferred ARGs, and even when defining
what the criteria for success should be.
It also means there is no shared format for interchanging ARG data between
programs, resulting in the assorted negative consequences with which
geneticists are only too familiar~\citep{excoffier2006computer,gower2022demes}.
Perhaps most importantly, the direct linkage to
the original stochastic process definition leads to an
overly narrow interpretation of ARGs, which is significantly
out of step with modern methods and large-scale datasets, and
can lead to missed opportunities in their application and usage.
Such problems go much deeper than pedantic discussion over terminology;
they directly inform what we should be attempting to estimate, how we
assess success, and the expected limitations on the precision
and accuracy of our inferences.
% Thanks Gideon and Alex!
If not addressed, these problems will
impede progress in applying genetic genealogies to understand the biological world.

@@

Fig.~\ref{fig-ancestry-resolution}C shows the graph that we obtain
by resolving the ancestral material transmitted along each edge
using the ``simplify'' algorithm~\citep{kelleher2018efficient},
and highlights some important
differences between the eARG and gARG encodings.
Firstly, we can see that some nodes and edges are removed entirely
from the graph by this conversion process.
The ``grand MRCA'' \noderef{q} is omitted from the
sample-resolved gARG because all segments of the genome have
fully coalesced before it is reached. Likewise, the edge
between \noderef{g} and \noderef{j} is omitted because the recombination
event at position $5$ (represented by node \noderef{g})
fell in non-ancestral material.
More generally, we can see that the sample resolved
retrospective gARG of Fig.~\ref{fig-ancestry-resolution}C
allows for ``local'' inspection
of an ARG in a way that is not possible by storing
common ancestor and recombination events in an eARG. Because
the ancestral material is stored with each edge, the
cumulative effects of events over time can be reasoned
about, without first ``replaying'' those events. Many computations
that we wish to perform on an ARG will require resolving
the ancestral material with respect to a sample.
% this is weak, but would be good to emphasise this point.
% you just end up running simplify lots of times, in your
% downstream algs
The gARG encoding
allows us to perform this resolution of ancestral material once 
and to store the result, 
whereas the eARG encoding requires us to repeat the process 
each time.

@@

The gARG encoding discussed and concretely defined in this manuscript
leads to highly efficient storage and processing of ARG data,
and has already been in use for several years.
The succinct tree sequence data structure (usually known as a ``tree sequence''
for brevity)
is a concrete encoding of a gARG, as discussed in this article.
It was originally developed as part of the \texttt{msprime}
simulator~\citep{kelleher2016efficient} and has subsequently been
extended and applied to forward-time
simulations~\citep{kelleher2018efficient,haller2018tree},
inference from data~\citep{kelleher2019inferring,wohns2022unified},
and calculation of population genetics statistics~\citep{ralph2020efficiently}.
The succinct tree sequence encoding extends the basic definition
of a gARG provided here by stipulating a
simple tabular representation and nodes and edges,
and also defining a concise and lossless representation of
sequence variation using the ``site'' and  ``mutation'' tables.
(Technically, edges in tree sequence terminology would be better
described as ``edge-intervals'', as each describes a single contiguous
interval of genome inheritance between a pair of nodes. This
denormalisation of the gARG data model in \texttt{tskit} is for efficiency purposes.)
The \texttt{tskit} library is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with gARGs (encoded as a succinct tree sequence).
Based on core functionality written
in C, it provides interfaces in C, Python and Rust.
Tskit is mature software, widely used in population genetics, and
has been incorporated into several downstream
applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
adrion2020community,
terasaki2021geonomics,
baumdicker2021efficient,
fan2022genealogical,korfmann2022weak,
mahmoudi2022bayesian,petr2022slendr,rasmussen2022espalier,
zhang2023biobank,nowbandegani2023extremely,ignatieva2023distribution}.
